### 2024-02-28 Introduction.tex
Fixed minitocs

\chapter{Introduction}
\label{chap:intro}
\minitoc

@@

This chapter presents some clinical context of brain tumours (Section \ref{sec:braintumours}) and aneurysms (Section \ref{sec:aneurysms}) for which this research is targeted alongside the current, relevant, available clinical adjuncts. This is followed by a description of oxygen saturation (Section \ref{sec:oxygensat}) and optical properties of tissues (Section \ref{sec:introoptics} alongside some key methods to calculate oxygen saturation using optical measurements. This is followed by some key existing technologies used for measurement of tissue and their limitations (Section \ref{sec:stateofart}). Finally, the structure of this thesis is presented in Section \ref{sec:thesisoutline}.

@@

\subsection{Categorising tumours} 
Tumours can fall into one of the two categories: malignant or benign. A malignant tumour can invade local tissues or some cells can travel through the blood stream and cause tumours in other regions of the body, labelled metastatic tumours. In contrast a benign tumour will not invade local or distal tissues, and, unlike malignant tumours, are likely not to grow back if removed~\citep{Institute2021}. The likelihood of a tumour to grow and spread determines its grade, which can range from Grade 1 (rarely spread and can generally can be cured by surgery) to Grade 4 (grow and spread very quickly and usually cannot be cured)~\citep{Institute2023}. Unlike other regions of the body, even benign tumours in the brain can be life-threatening. A brain tumour can grow and press on regions of the brain causing those areas to stop working leading to a range of symptoms. This is true for both primary brain tumours, that originate in the brain, and metastatic brain tumours, that originate elsewhere in the body~\citep{Institute2021}.

The type of tumour and it's impact is determined by the region of brain it affects. There are three main sections of the brain which each control different aspects of human functioning. Some examples are shown in Figure \ref{fig:anatomybrain} and include the cerebrum, which is the largest part of the brain and controls thinking, learning, problem solving, emotions, speech, reading, writing, and voluntary movement; the cerebellum controls movement, balance, and posture; and the brain stem which connects the brain to the spinal cord and controls breathing, heart rate, and the nerves and muscles used to see, hear, walk, talk, and eat~\citep{Institute2023}.
\begin{figure}[ht]
    \centering
    \includegraphics[width=0.75\textwidth]{brainanatomy.jpg}
    \caption{Depiction of some key regions of the brain reproduced from~\citet{Institute2023}.}
    \label{fig:anatomybrain}
\end{figure}
There are also different types of brain cells in which the tumour can originate. The brain contains many neurones which transport the electrical signals that allow the brain to operate. These are surrounded by glial cells, of which there are different types which each play a role in supporting and protecting the neurones~\citep{TheBrainTumourCharity2023}. The main types are astrocytes, oligodendrocytes and ependymal cells. When the tumour originates in glial cells it is termed a glioma with subcategories corresponding to each type of glial cell. Astrocytomas, originating in astrocytes, are the most common gliomas with the different grades given different names such as Grade 4 astrocytoma named glioblastoma multiforme~\citep{BrainTumourResearch2023}. Similarly, oligodendrogliomas originate from oligodendrocytes, and ependymomas originate from ependymal cells~\citep{TheBrainTumourCharity2023}. %include meningioma lymphoma medulloblastoma adenoma \\

@@

\subsection{Current treatments}\label{sec:introtumourtreatments}
Initially or if no other treatment options are possible, the symptoms of a brain tumour can be treated without treating the tumour itself, for example anti-sickness medication to treat the nausea. The treatment plans are determined based on a number of factors which include the type, location, and grade of the tumour as well as the patient's overall health~\citep{NationalHealthService2023}. Surgery is a preferred treatment where possible as it is possible to remove a large proportion, if not all, of the tumour with this method. In cases where the full tumour may not be removed, such as when there is a significant infiltration zone into surrounding healthy tissue where the border of the tumour is challenging to identify, this may be followed up with radiotherapy or chemotherapy. These treatment options can also be used separately when surgery is not an option~\citep{MacmillanCancerSupport2019}.

Neurosurgical resection aims for maximum safe resection of the tumour which balances a high extent of resection while preserving functional regions of brain as improved resection is linked to improved patient outcomes~\citep{Chanbour2022}. To this end a variety of techniques can be employed to improve resection. These include: a surgical microscope which enables clear visualisation of the surgical field, neuronavigation which allows registration of a probe against an MRI, awake or asleep sub-corticol mapping to allow identification of eloquent areas of brain, in some cases fluorescence imaging using 5-aminolevulinic acid (5-ALA) or indocyanine green (ICG) can be used for tumour localisation, and microvascular doppler (MVD) or ICG can be used for tumour vessel localisation~\citep{Chanbour2022, Catapano2018}.

@@

\section{Brain aneurysms}\label{sec:aneurysms}
Another key set of conditions addressed by neurosurgery include brain aneurysms. These are a subset of all aneurysms which are balloon-like abnormal dilatations in a vessel wall due to a weakness caused by high blood pressure, smoking, or genetic predisposition~\citep{NationalHealthService2022}. Most aneurysms will only have significant symptoms on rupturing which is called a subarachnoid haemmorhage and can cause extensive brain damage as well as blinding headaches, sickness, or pain looking at light~\citep{NationalHealthService2022} with approximately 500,000 people dying annually~\citep{Toth2018}. To aid the precise nature of aneurysm surgical treatment, we aim to aid surgical guidance by providing more information to the surgeon using hyperspectral imaging.

\subsection{Current treatments}\label{sec:introaneurysmtreatments}
Treatment of unruptured aneurysm is determined by risk of rupture, size, and location in the brain. If the aneurysm is considered to have a low risk of rupture, treatment may focus on limiting risk factors~\citep{NationalHealthService2022}. In ruptured aneurysms or those at higher risk of rupturing, however, treatment is primarily surgical in the form of clipping or endovascular procedures depicted in Figure \ref{fig:aneurysmtreatment}. The former involves the placement of a metal clip over the opening to prevent further blood flow into the aneurysm~\citep{TheBrainFoundation2023}. Endovascular procedures can include coiling to encourage clotting in the aneurysm to prevent further blood flow, which can be combined with further devices including stents to hold the coil in place~\citep{TheBrainFoundation2023}. The final endovascular technique is flow diversion achieved by placing a cylindrical tube (similar to a stent with a tighter weave) in the vessel over the aneurysm opening to allow normal blood flow~\citep{TheBrainFoundation2023}. 
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{Aneurysm-treatments.png}
    \caption{An illustration of some key surgical treatment methods for brain aneurysms reproduced from~\citet{TheBrainFoundation2023} where the technique depicted by A is coiling, B is clipping, C is coiling combined with a stent, and D is flow diversion.}
    \label{fig:aneurysmtreatment}
\end{figure}
Whilst these treatment methods are often very effective, incomplete aneurysm occlusion can lead to rupture at a later date with the extent of occlusion being linked to the likelihood of this occurring~\citep{Toth2018}. For this reason blood flow is often assessed intra-operatively using digital subtraction angiography (DSA), MVD, or ICG~\citep{Norat2019}. DSA is produced by subtracting an X-ray image taken without a contrast agent from one taken with a fluorescent contrast agent to visualise only the vessels in the field of view~\citep{Radiopia2022}. This can be done intra-operatively to visualise the blood flow, however the gold-standard for determining quality of aneurysm occlusion is a post-operative DSA~\citep{Marbacher2020}. DSA cannot be used continuously intra-operatively due to the changing geometry, the complex procedure for DSA aquisition, and the radiation dose risk~\citep{Radiopia2022, Derdeyn1999}. ICG provides a real-time alternative to DSA, however it also requires a fluorescent dye and is of lower precision~\citep{Norat2019, Anania2023}. MVD is also able to determine if full occlusion has occurred intra-operatively and has the advantage of not requiring a labelling agent, however this technique is highly subject to angle and depth~\citep{Anania2023}. Additionally, temporary clipping of healthy vessels can be used to control intra-operative bleeding, however this may lead to cerebral ischaemia where tissues have limited access to oxygenated blood~\citep{Doron2022}.

Similarly to those in Section \ref{sec:introtumourtreatments}, the advantages of the current clinical adjuncts provide valuable information of the desirable features in technology to aid aneurysm treatment. DSA, ICG and MVD all demonstrate the desirability of spatially-resolved evaluation technique. DSA and MVD highlight the importance of accounting for changing geometry or imaging position, while ICG draws attention to the advantage of label-free techniques. While ICG can distinguish between arterial and venous vessels based on their visualisation time, there are also no clinically available techniques to determine tissue oxygenation status. For this reason a real-time, intra-operative, spatially-resolved, quantitative measure of oxygen saturation which can be applied to both vessels and tissues in a range of imaging geometries is desired. HSI has been shown to address many of these challenges with steps towards the geometry-invariant imaging being addressed in this work.

@@

Oxygen is required by all cells in the body to release energy by respiration. This oxygen is transported to cells by haemoglobin, primarily in the red blood cells of blood, which can take the forms of oxyhaemoglobin and deoxyhaemoglobin based on whether the haemoglobin is bound to oxygen or not respectively. These forms of haemoglobin have very different absorbance characteristics as seen in their extinction spectra in Figure \ref{fig:Haemoglobinext}. 
\begin{figure}[h]
    \centering 
    \includegraphics[width=0.7\textwidth]{Hb_eps.pdf}
    \caption{The extinction coefficients for oxyhaemoglobin and deoxyhaemoglobin reproduced from~\citet{Prahl1998}.}
    \label{fig:Haemoglobinext}
\end{figure}
Oxygen saturation is defined as the proportion of haemoglobin which is oxygenated ie $\frac{HbO_2}{HbO_2 + Hb}$ where $Hb$ is deoxyhaemoglobin and $HbO_2$ is oxyhaemoglobin.

There are no currently established, intra-operative, techniques to assess $StO_2$ in neurosurgery and so perfusion is often used as an indicator instead with DSA, MVD, and ICG capable of this as detailed in Sections \ref{sec:introtumourtreatments} and \ref{sec:introaneurysmtreatments}. Additionally, there has been significant research into using Laser Speckle Contrast Imaging (LSCI) to visualise blood perfusion, however this is not able to distinguish between $Hb$ and $HbO_2$~\citep{Dunn2012, Zhong2021}. Peri-operative neural monitoring can use jugular venous oxygen saturation (JVOS)~\citep{Raith2020} and cerebral oximetry~\citep{Lian2020} to provide indications of the health of the organ. Both techniques usually make use of near infra-red (NIR) spectroscopy to distinguish between $Hb$ and $HbO_2$. These techniques, however, lack specificity so local regions of ischaemia can be present when these values are within the bounds of normality~\citep{Raith2020, Zhong2021}.

%Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from tissue diffuse reflectance spectra. These range from the simplest ratiometric two-wavelength methods to more sophisticated analytical models~\citep{MacKenzie2018} or deep-learning based approaches~\citep{Ayala2023}. Using two or three wavelength models can provide good results but require data to be captured at these precise wavelengths, whereas analytical models can be applied to ranges of wavelengths for which they are developed. These analytical models also enable more detailed modelling of tissue including consideration of melanin when imaging through skin. Alternatively, Monte Carlo methods are an established approach to modelling light-tissue interaction for a range of wavelengths. It is, however, computationally challenging to apply in an inverse problem setting making the estimation of optical properties from tissue spectra non trivial on the basis of Monte Carlo simulations alone. To obtain the diffuse reflectance spectra intra-operatively, these techniques primarily use HSI which enable non-contact imaging with minimal change to surgical workflow. 
Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from optical measurements of tissues, including diffuse reflectance spectra. These models relate the extinction coefficients of haemoglobin to the optical properties of tissues to enable $StO_2$ calculation.

\section{Optical properties of tissues}\label{sec:introoptics}
To extract useful information, such as $StO_2$, from optical measurements of tissues it is necessary to be able to model the optical properties of these tissues. The magnitude of the absorption and scattering of light through tissue is characterised by an absorption coefficient $\mu_a$ and scattering coefficient $\mu_s$. Scattering can occur isotropically, the light wave has an equal probability of propagating in any direction from the point of scatter, or anisotropically, where the light wave is more likely to propogate in a direction either forwards or backwards from the point of scatter~\citep{Jacques2013}. This directional component of scatter is described by the anisotropy factor $g$ for bulk tissue, where 0 represents isotropic scatter and 1 or -1 represents perfect forwards or backwards scattering respectively~\citep{Jacques2013}. These scattering parameters are often combined into a single scattering measure $\mu_s'$ termed the reduced scattering coefficient according to Equation \eqref{eq:reducedRintro}~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = \mu_s(1-g)
    \label{eq:reducedRintro}
\end{equation}

$\mu_a$ is well defined by combining the concentrations ($c$) and extinction coefficients ($\epsilon$) of each chromophore ($i$) or combining the volume fraction ($f$) and the absorption coefficient ($\mu_a$) of each pure chromophore ($i$) in the sample at any wavelength ($\lambda$)~\citep{Jacques2013, MacKenzie2018}. 
\begin{equation}
    \mu_a(\lambda) = \sum_i c_i \epsilon_i(\lambda) = \sum_i f_i \mu_{a, i}(\lambda)
    \label{eq:definemua}
\end{equation}
For non-scattering media this $\mu_a$ can easily be converted to an absorbance measurement ($A$) by multiplying with the path length ($l$) of the sample through which the measurement is obtained according to the Beer-Lambert law~\cite{MacKenzie2018}. This also allows simple calculations to solve the inverse problem of calculating concentrations of chromophores from absorbance measurements of samples. 
\begin{equation}
    A = \mu_al
\end{equation}

Scattering behaviour of particles in biological tissues can be broadly divided into Rayleigh, Mie, and Raman scattering. Commonly, Rayleigh and Mie scattering are defined as the elastic scattering behaviour by particles significantly smaller or particles comparable to or larger than the incident wavelength of light respectively. More precisely, Mie scattering refers to the scattering of spheres of any size with Rayleigh scattering being the Rayleigh limit of Mie scattering for particles significantly smaller than the wavelength of incident light~\cite{Jacques2013}. Raman scattering refers to the inelastic scattering of light, however this effect is typically much smaller and is not easy to detect without long integration times for typical optical measurement systems~\citep{Kong2015}. Within the common measurement range of 400-1000nm, modelling Mie scattering alone is sufficient, however beyond this range a combination of Rayleigh and Mie scattering may be required to provide a good model of the scattering behaviour~\cite{Jacques2013}. This Rayleigh component at extreme wavelengths can be more significant for tissues with higher levels of collagen, whereas Mie theory continues to provide a good model of scattering for tissues where the predominant scatterer is fat~\cite{Jacques2013}. By using Mie theory, each of the scattering coefficients can be calculated using the diameter and volume fraction of the scattering particles for each wavelength of light~\citep{Hergert2012}. This information, however, is usually unknown for biological tissue and so the Mie scattering is often simplified to Equation \eqref{eq:Mieintro} where the reduced scattering at 500nm is taken as a reference point~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = a(\frac{\lambda}{500nm})^{-b}
    \label{eq:Mieintro}
\end{equation}

\subsection{Major $StO_2$ approaches}
By considering these optical properties of tissues, a variety of methods to extract $StO_2$ from measurements of tissue have been developed. In this section we will discuss some of the key methods used to extract $StO_2$ when using the range of wavelengths 400-1000nm to obtain in vivo measurements.

The most simplistic method of $StO_2$ estimation was the two-wavelength approach based on a pure Beer-Lambert model of tissue. This method assumes only oxygenated and deoxygenated haemoglobin ($HbO_2$ and $Hb$) as chromophores in tissue. 
\begin{equation}
    \mu_a = c_{Hb}\epsilon_{Hb} + c_{HbO_2}\epsilon_{HbO_2}
\end{equation}
In this method measurements of tissue are acquired at two wavelengths, one of which is an isosbestic point for the haemoglobin chromophore where the oxygenated and deoxygenated extinction coefficients are equal~\citep{MacKenzie2018}. This approach required significant calibration to account for scattering in tissues~\citep{MacKenzie2018}. Pulse oximetry is based on this approach, however the pulsatile nature of arterial blood allows the background non-pulsatile absorbance to be removed meaning that arterial $SpO_2$ can be calculated for the pulsatile portion of the signal only~\citep{MacKenzie2018}. This, however, is still confounded by the presence of other significant absorbers in high concentrations such as melanin or by low blood flow due to low temperature~\citep{Bangash2022, MacKenzie2018}. This allowed development of the three-wavelength approach. In this case, two isosbestic wavelengths are measured to calculate a wavelength-independent scattering coefficient with a third used for $StO_2$ extraction~\cite{MacKenzie2018}. Whilst this provides advantages over the two-wavelength approach, it has largely been superseded by hyperspectral approaches due to the limited number of appropriate wavelengths that can be used which satisfy the approximation of constant scattering and by the continued requirement for calibration with samples of known $StO_2$~\cite{MacKenzie2018}. By using ranges of wavelengths and investigating the derivative of the spectrum a hyperspectral alternative has been proposed, however this also assumes wavelength-independent scattering and requires significant calibration~\citep{Holmer2018}. To advance this field, Monte Carlo simulations were used to probabilistically predict the spectra that could be obtained from a sample with known optical properties~\cite{Wang1995}. These have been used to improve approaches based on the Beer-Lambert models by including differential path length factors which are fitted to simulated data~\cite{Clancy2020}.

>>>>

\chapter{Introduction}
\label{chap:intro}

\minitoc

@@

This chapter presents some clinical context of brain tumours (Section~\ref{sec:braintumours}) and aneurysms (Section~\ref{sec:aneurysms}) for which this research is targeted alongside the current, relevant, available clinical adjuncts. This is followed by a description of oxygen saturation (Section~\ref{sec:oxygensat}) and optical properties of tissues (Section~\ref{sec:introoptics}) alongside some key methods to calculate oxygen saturation using optical measurements. This is followed by some key existing technologies used for measurement of tissue and their limitations (Section~\ref{sec:stateofart}). Finally, the structure of this thesis is presented in Section~\ref{sec:thesisoutline}.

@@

\subsection{Categorising tumours} 
Tumours can fall into one of two categories: malignant or benign. A malignant tumour can invade local tissues or some cells can travel through the blood stream and cause tumours in other regions of the body, labelled metastatic tumours. In contrast a benign tumour will not invade local or distal tissues, and, unlike malignant tumours, are likely not to grow back if removed~\citep{Institute2021}. The likelihood of a tumour to grow and spread determines its grade, which can range from Grade 1 (rarely spread and can generally be cured by surgery) to Grade 4 (grow and spread very quickly and usually cannot be cured)~\citep{Institute2023}. Unlike other regions of the body, even benign tumours in the brain can be life-threatening. A brain tumour can press on regions of the brain causing those areas to stop working and leading to a range of symptoms. This is true for both primary brain tumours, that originate in the brain, and metastatic brain tumours, that originate elsewhere in the body~\citep{Institute2021}.

The type of tumour and it's impact is determined by the region of brain it affects. There are three main sections of the brain which each control different aspects of human functioning. Some examples are shown in Figure~\ref{fig:anatomybrain} and include the cerebrum, which is the largest part of the brain and controls thinking, learning, problem solving, emotions, speech, reading, writing, and voluntary movement; the cerebellum controls movement, balance, and posture; and the brain stem which connects the brain to the spinal cord and controls breathing, heart rate, and the nerves and muscles used to see, hear, walk, talk, and eat~\citep{Institute2023}.
\begin{figure}[ht]
    \centering
    \includegraphics[width=0.75\textwidth]{brainanatomy.jpg}
    \caption{Depiction of some key regions of the brain reproduced from~\citep{Institute2023}.}
    \label{fig:anatomybrain}
\end{figure}
There are also different types of brain cells in which the tumour can originate. The brain contains many neurones which transport the electrical signals that allow the brain to operate. These are surrounded by glial cells, of which there are different types which each play a role in supporting and protecting the neurones~\citep{TheBrainTumourCharity2023}. The main types are astrocytes, oligodendrocytes and ependymal cells. When the tumour originates in glial cells it is termed a glioma with subcategories corresponding to each type of glial cell. Astrocytomas, originating in astrocytes, are the most common gliomas with the different grades given different names such as Grade 4 astrocytoma named glioblastoma multiforme~\citep{BrainTumourResearch2023}. Similarly, oligodendrogliomas originate from oligodendrocytes, and ependymomas originate from ependymal cells~\citep{TheBrainTumourCharity2023}. %include meningioma lymphoma medulloblastoma adenoma \\

@@

\subsection{Current treatments}\label{sec:introtumourtreatments}
Initially, or if no other treatment options are possible, the symptoms of a brain tumour can be treated without treating the tumour itself, for example anti-sickness medication to treat the nausea. The treatment plans are determined based on a number of factors which include the type, location, and grade of the tumour as well as the patient's overall health~\citep{NationalHealthService2023}. Surgery is a preferred treatment where possible as it is possible to remove a large proportion, if not all, of the tumour with this method. In cases where the full tumour may not be removed, such as when there is a significant infiltration zone into surrounding healthy tissue meaning the border of the tumour is challenging to identify, this may be followed with radiotherapy or chemotherapy. These treatment options can also be used separately when surgery is not an option~\citep{MacmillanCancerSupport2019}.

Neurosurgical resection aims for maximal safe resection of the tumour which balances a high extent of resection while preserving functional regions of brain as improved resection is linked to improved patient outcomes~\citep{Chanbour2022}. To this end, a variety of techniques can be employed to improve resection. These include: a surgical microscope which enables clear visualisation of the surgical field, neuronavigation which allows registration of a probe against an MRI scan, awake or asleep sub-corticol mapping to allow identification of eloquent areas of brain, in some cases fluorescence imaging using 5-aminolevulinic acid (5-ALA) or indocyanine green (ICG) can be used for tumour localisation, and microvascular doppler (MVD) or ICG can be used for tumour vessel localisation~\citep{Chanbour2022, Catapano2018}.

@@

\section{Brain aneurysms}\label{sec:aneurysms}
Another key set of conditions addressed by neurosurgery include brain aneurysms. These are a subset of all aneurysms which are balloon-like abnormal dilatations in a vessel wall due to a weakness caused by high blood pressure, smoking, or genetic predisposition~\citep{NationalHealthService2022}. Most aneurysms will only have significant symptoms on rupturing which is called a subarachnoid haemmorhage and can cause extensive brain damage as well as blinding headaches, sickness, or pain looking at light~\citep{NationalHealthService2022} with approximately 500,000 people dying annually~\citep{Toth2018}. To aid the precise nature of surgical treatment of aneurysms, we aim to aid surgical guidance by providing more information to the surgeon using hyperspectral imaging.

\subsection{Current treatments}\label{sec:introaneurysmtreatments}
Treatment of unruptured aneurysm is determined by the risk of rupture, size, and location in the brain. If the aneurysm is considered to have a low risk of rupture, treatment may focus on limiting risk factors~\citep{NationalHealthService2022}. In ruptured aneurysms or those at higher risk of rupturing, however, treatment is primarily surgical in the form of clipping or endovascular procedures depicted in Figure~\ref{fig:aneurysmtreatment}. The former involves the placement of a metal clip over the opening to prevent further blood flow into the aneurysm~\citep{TheBrainFoundation2023}. Endovascular procedures can include coiling to encourage clotting in the aneurysm to prevent further blood flow, which can be combined with further devices including stents to hold the coil in place~\citep{TheBrainFoundation2023}. The final endovascular technique is flow diversion achieved by placing a cylindrical tube (similar to a stent with a tighter weave) in the vessel over the aneurysm opening to allow normal blood flow~\citep{TheBrainFoundation2023}. 
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{Aneurysm-treatments.png}
    \caption{An illustration of some key surgical treatment methods for brain aneurysms reproduced from~\citep{TheBrainFoundation2023} where the technique depicted by A is coiling, B is clipping, C is coiling combined with a stent, and D is flow diversion.}
    \label{fig:aneurysmtreatment}
\end{figure}

Whilst these treatment methods are often very effective, incomplete aneurysm occlusion can lead to rupture at a later date with the extent of occlusion being linked to the likelihood of this occurring~\citep{Toth2018}. For this reason, blood flow is often assessed intra-operatively using digital subtraction angiography (DSA), MVD, or ICG~\citep{Norat2019}. DSA is produced by subtracting an X-ray image taken without a contrast agent from one taken with a fluorescent contrast agent to visualise only the vessels in the field of view~\citep{Radiopia2022}. This can be done intra-operatively to visualise the blood flow, however the gold-standard for determining quality of aneurysm occlusion is post-operative DSA~\citep{Marbacher2020}. DSA cannot be used continuously intra-operatively due to the changing geometry, the complex procedure for DSA aquisition, and the radiation dose risk~\citep{Radiopia2022, Derdeyn1999}. ICG provides a real-time alternative to DSA, however it also requires a fluorescent dye and is of lower precision~\citep{Norat2019, Anania2023}. MVD is also able to determine if full occlusion has occurred intra-operatively and has the advantage of not requiring a labelling agent, however this technique is highly subject to angle and depth~\citep{Anania2023}. Additionally, temporary clipping of healthy vessels can be used to control intra-operative bleeding, however this may lead to cerebral ischaemia where tissues have limited access to oxygenated blood~\citep{Doron2022}. Currently, however, there are no widely available methods of assessing tissue oxygenation directly leading to subjective assessment of tissue viability.

Similarly to those in Section~\ref{sec:introtumourtreatments}, the advantages of the current clinical adjuncts provide valuable information of the desirable features in technology developed to aid aneurysm treatment. DSA, ICG and MVD all demonstrate the value of spatially-resolved evaluation techniques. DSA and MVD highlight the importance of accounting for changing geometry or imaging position, while ICG draws attention to the advantage of label-free techniques. While ICG can distinguish between arterial and venous vessels based on their visualisation time, there are also no clinically available techniques to determine tissue oxygenation status. For this reason a real-time, intra-operative, spatially-resolved, quantitative measure of oxygen saturation which can be applied to both vessels and tissues in a range of imaging geometries is desired. HSI has been shown to address many of these challenges with steps towards the geometry-invariant imaging being addressed in this work.

@@

Oxygen is required by all cells in the body to enable release of energy by respiration. This oxygen is transported to cells by haemoglobin, primarily in the red blood cells of blood, which can take the forms of oxyhaemoglobin and deoxyhaemoglobin based on whether the haemoglobin is bound to oxygen or not respectively. These forms of haemoglobin have very different absorbance characteristics as seen in their extinction spectra in Figure~\ref{fig:Haemoglobinext}. 
\begin{figure}[h]
    \centering 
    \includegraphics[width=0.7\textwidth]{Hb_eps.pdf}
    \caption{The extinction coefficients for oxyhaemoglobin and deoxyhaemoglobin reproduced from~\citep{Prahl1998}.}
    \label{fig:Haemoglobinext}
\end{figure}
Oxygen saturation is defined as the proportion of haemoglobin which is oxygenated i.e. $\frac{HbO_2}{HbO_2 + Hb}$ where $Hb$ is deoxyhaemoglobin and $HbO_2$ is oxyhaemoglobin.

Currently, there are no established, intra-operative, techniques to assess $StO_2$ in neurosurgery and so perfusion is often used as an indicator instead with DSA, MVD, and ICG capable of this as detailed in Sections~\ref{sec:introtumourtreatments} and~\ref{sec:introaneurysmtreatments}. Additionally, there has been significant research into using Laser Speckle Contrast Imaging (LSCI) to visualise blood perfusion, however this is not able to distinguish between $Hb$ and $HbO_2$~\citep{Dunn2012, Zhong2021}. Peri-operative neural monitoring can use jugular venous oxygen saturation (JVOS)~\citep{Raith2020} and cerebral oximetry~\citep{Lian2020} to provide indications of the health of the organ. Both techniques usually make use of near infra-red (NIR) spectroscopy to distinguish between $Hb$ and $HbO_2$. These techniques, however, lack specificity so it is possible for local regions of ischaemia to be present even when these values are within the bounds of normality~\citep{Raith2020, Zhong2021}.

%Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from tissue diffuse reflectance spectra. These range from the simplest ratiometric two-wavelength methods to more sophisticated analytical models~\citep{MacKenzie2018} or deep-learning based approaches~\citep{Ayala2023}. Using two or three wavelength models can provide good results but require data to be captured at these precise wavelengths, whereas analytical models can be applied to ranges of wavelengths for which they are developed. These analytical models also enable more detailed modelling of tissue including consideration of melanin when imaging through skin. Alternatively, Monte Carlo methods are an established approach to modelling light-tissue interaction for a range of wavelengths. It is, however, computationally challenging to apply in an inverse problem setting making the estimation of optical properties from tissue spectra non trivial on the basis of Monte Carlo simulations alone. To obtain the diffuse reflectance spectra intra-operatively, these techniques primarily use HSI which enable non-contact imaging with minimal change to surgical workflow. 
Due to this lack of available techniques, there have been many proposed methods to extract $StO_2$ from optical measurements of tissues, including from diffuse reflectance spectra. These models relate the extinction coefficients of haemoglobin to the optical properties of tissues to enable $StO_2$ calculation.

\section{Optical properties of tissues}\label{sec:introoptics}
To extract useful information, such as $StO_2$, from optical measurements of tissues it is necessary to be able to model the optical properties of these tissues. The magnitude of the absorption and scattering of light through tissue is characterised by an absorption coefficient $\mu_a$ and scattering coefficient $\mu_s$. Scattering can occur isotropically, the light wave has an equal probability of propagating in any direction from the point of scatter, or anisotropically, where the light wave is more likely to propogate in a direction either forwards or backwards from the point of scatter~\citep{Jacques2013}. This directional component of scatter is described by the anisotropy factor $g$ for bulk tissue, where 0 represents isotropic scatter and 1 or -1 represents perfect forwards or backwards scattering respectively~\citep{Jacques2013}. These scattering parameters are often combined into a single scattering measure $\mu_s'$ termed the reduced scattering coefficient according to Equation~\eqref{eq:reducedRintro}~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = \mu_s(1-g)
    \label{eq:reducedRintro}
\end{equation}

$\mu_a$ is well defined by combining the concentrations ($c$) and extinction coefficients ($\epsilon$) of each chromophore ($i$) or combining the volume fraction ($f$) and the absorption coefficient ($\mu_a$) of each pure chromophore ($i$) in the sample at any wavelength ($\lambda$)~\citep{Jacques2013, MacKenzie2018}. 
\begin{equation}
    \mu_a(\lambda) = \sum_i c_i \epsilon_i(\lambda) = \sum_i f_i \mu_{a, i}(\lambda)
    \label{eq:definemua}
\end{equation}
For non-scattering media this $\mu_a$ can easily be converted to an absorbance measurement ($A$) by multiplication with the path length ($l$) through which the measurement is obtained according to the Beer-Lambert law~\citep{MacKenzie2018}. This also allows simple calculations to solve the inverse problem of calculating concentrations of chromophores from absorbance measurements of samples. 
\begin{equation}
    A = \mu_al
\end{equation}

Scattering behaviour of particles in biological tissues can be broadly divided into Rayleigh, Mie, and Raman scattering. Commonly, Rayleigh and Mie scattering are defined as the elastic scattering behaviour by particles significantly smaller or particles comparable to or larger than the incident wavelength of light respectively. More precisely, Mie scattering refers to the scattering of spheres of any size with Rayleigh scattering being the Rayleigh limit of Mie scattering for particles significantly smaller than the wavelength of incident light~\citep{Jacques2013}. Raman scattering refers to the inelastic scattering of light, however this effect is typically much smaller and is not easy to detect without long integration times for typical optical measurement systems~\citep{Kong2015}. Within the common measurement range of 400-1000nm, modelling Mie scattering alone is sufficient, however beyond this range a combination of Rayleigh and Mie scattering may be required to provide a good model of the scattering behaviour~\citep{Jacques2013}. This Rayleigh component at extreme wavelengths can be more significant for tissues with higher levels of collagen, whereas Mie theory continues to provide a good model of scattering for tissues where the predominant scatterer is fat~\citep{Jacques2013}. By using Mie theory, each of the scattering coefficients can be calculated using the diameter and volume fraction of the scattering particles for each wavelength of light~\citep{Hergert2012}. This information, however, is usually unknown for biological tissue and so the Mie scattering is often simplified to Equation~\eqref{eq:Mieintro} where the reduced scattering at 500nm is taken as a reference point and scattering coefficients $a$ and $b$ vary across tissues~\citep{Jacques2013}. 
\begin{equation}
    \mu_s' = a(\frac{\lambda}{500nm})^{-b}
    \label{eq:Mieintro}
\end{equation}
By considering these optical properties of tissues, a variety of methods to extract $StO_2$ from measurements of tissue have been developed.

\subsection{Major $StO_2$ approaches}
In this section we will discuss some of the key methods used to extract $StO_2$ when using the range of wavelengths 400-1000nm to obtain in vivo measurements.

The most simplistic method of $StO_2$ estimation was the two-wavelength approach based on a pure Beer-Lambert model of tissue. This method assumes only oxygenated and deoxygenated haemoglobin ($HbO_2$ and $Hb$) as chromophores in tissue. 
\begin{equation}
    \mu_a = c_{Hb}\epsilon_{Hb} + c_{HbO_2}\epsilon_{HbO_2}
\end{equation}
In this method measurements of tissue are acquired at two wavelengths, one of which is an isosbestic point for the haemoglobin chromophore where the oxygenated and deoxygenated extinction coefficients are equal so the measurement is invariant to $StO_2$~\citep{MacKenzie2018}. This approach required significant calibration to account for scattering in tissues~\citep{MacKenzie2018}. Pulse oximetry is based on this, however the pulsatile nature of arterial blood allows the background non-pulsatile absorbance to be removed meaning that arterial $SpO_2$ can be calculated for the pulsatile portion of the signal only~\citep{MacKenzie2018}. This, however, is still confounded by the presence of other significant absorbers in high concentrations such as melanin or by low blood flow due to low temperature~\citep{Bangash2022, MacKenzie2018}. This allowed development of the three-wavelength approach. In this case, two isosbestic wavelengths are measured to calculate a wavelength-independent scattering coefficient with a third used for $StO_2$ extraction~\citep{MacKenzie2018}. Whilst this provides advantages over the two-wavelength approach, it has largely been superseded by hyperspectral approaches due to the limited number of appropriate wavelengths that can be used which satisfy the approximation of constant scattering and by the continued requirement for calibration with samples of known $StO_2$~\citep{MacKenzie2018}. By using ranges of wavelengths and investigating the derivative of the spectrum a hyperspectral alternative has been proposed, however this also assumes wavelength-independent scattering and requires significant calibration~\citep{Holmer2018}. To advance this field, Monte Carlo simulations were used to probabilistically predict the spectra that could be obtained from a sample with known optical properties~\citep{Wang1995}. These have been used to improve approaches based on the Beer-Lambert models by including differential path length factors which are fitted to simulated data~\citep{Clancy2020}.

Whilst Monte Carlo simulations provide reliable predictions of diffuse reflectance spectra, the inverse solution is computationally challenging which makes the estimation of optical properties from measured spectra non-trivial on the basis of Monte Carlo simulations alone. To make use of this, Monte Carlo based look-up tables have been used in some cases~\citep{Nguyen2021, Yang2022}. Alternatively, Monte Carlo is often used to validate or inform more complex analytical models such as those investigated in this work by \citet{Jacques1999} and \citet{Yudovsky2009} which are discussed in detail in later chapters. Finally, deep learning algorithms have been trained using Monte Carlo simulated data and applied in vivo~\citep{Wirkert2017, Wirkert2016}. Whilst these algorithms hold promise for improving the accuracy of $StO_2$ extraction, they rely on the accurate modelling of the input $\mu_a$ and $\mu_s$ parameters for the Monte Carlo simulations and often do not show validation with measured data of known ground-truth. As an initial step to understanding and improving these approaches, some analytical models are investigated further in this work.
